---
title: "SIR_closed_qrt"
author: "Kyu Lee"
date: "2023-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggplot2)
library(reshape2)
library(dplyr)
```

## Adds quarantine to the basic SIR model (closed population)
This code is to illustrate how to model quarantine in an SIR model without demography (no births or deaths). In the differential equation, we have Q compartment that indicates individuals who are infected and isolated under quarantine.

```{R}
# SIR model with quarantine
SIR_Qrtn<-function(t, state, params) {
  with(as.list(c(state, params)),{
    N = S + I + Q + R

    #SIR with quarantine
    dS<- -beta*S*I/N
    dI <- beta*S*I/N - q*I - gamma*I
    dQ <- q*I - gamma*Q
    dR <- gamma*I + gamma*Q
    dC <- beta*S*I/N

    # return the rates of change as a list
    list(c(dS, dI, dQ, dR, dC))
  })
}
```

We then define parameters and starting compartment sizes for ode solver. q is quarantine rate (I->Q)
```{r}
# 2. Define parameters and starting compartment sizes
params <- c(beta = 0.5, #effective contact rate (aka transmission rate)
                gamma = 0.3, #recovery rate (1/duration infection)
                q = 0.02 # quarantine rate
)
# Initial state
state.q <- c(S = 99999, #population of 100,000, 1 person starts of infected
             I = 1,
             Q = 0,
             R = 0,
             C = 0)
```

We set the total time steps and run the SIR model with quarantine
```{r}
T_end <- 500 #run model for 500 time steps (e.g. months)
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps (e.g. months), and computes output at each time step
# Run ODE solver
output.qrtn <- ode(y = state.q, times = times, func = SIR_Qrtn, parms = params)
```

Plot S,I,R,Q compartment
```{r, echo=FALSE}
# Plot base-case result
output.qrtn.t <- melt(as.data.frame(output.qrtn) %>% select(-C), id.vars="time")
ggplot(output.qrtn.t)+
  geom_line(aes(time,value,color=variable))+
  ylab("Number of individuals")+
  theme_bw()
```

Let's run a sensitivity analysis on quarantine rate to see how it affects the projected epidemic.
```{r}
# Sensitivity analysis on quarantine rate
q_list <- c(0,0.05,0.1) # a vector of quarantine rate
q_out_all <- data.frame() # an empty dataset to save the outcome

for(this_q in q_list){
  temp_param <- params #copy parameter
  temp_param[['q']] <- this_q # replace q
  # run ode solver
  this_out <- data.frame(ode(y = state.q, times = times, func = SIR_Qrtn, parms = temp_param))
  # save q value in this cycle
  this_out <- this_out %>%
    mutate(N = S+I+R+Q,
           q = this_q)
  # stack the result
  if(this_q == q_list[1]){
    q_out_all <- this_out
  }else{
    q_out_all <- rbind(q_out_all, this_out)
  }
}
# change the dataset to a long-form dataset
q_out_all_t <- melt(q_out_all%>% select(time,S,I,R,N,q), id.vars=c("time","q","N"))
# Plot S, I, R
ggplot(q_out_all_t) +
  geom_line(aes(x=time,y=value/N*100,color=variable))+
  ylab("Proportion of population")+
  facet_wrap(.~q)+
  theme_bw()
# Plot cumulative number of infections
ggplot(q_out_all) +
  geom_point(aes(x=time,y=C))+
  facet_wrap(.~q)+
  ylab("Cumulative infections")+
  theme_bw()
```
