---
title: "SI_contacttracing"
author: "Kyu Lee"
date: "2023-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggplot2)
library(reshape2)
library(dplyr)
```

## Adds vaccination to the basic SIR model
This code is to illustrate how to model contact tracing in SI model.

First, we defined the ode function with S,I,Tx compartment. Tx compartment indicates infected individuals who are under treatment and contact tracing. 
```{r}
# 1. Define model function
SI_CC<-function(t, state, parameters) {
  with(as.list(c(state, parameters)),{
    N = S + I + Tx
    
    #SI w/o demography equations from lecture
    dS <- -beta*S*I/N + rt*Tx
    dI <- beta*S*I/N - si*I - c*b*Tx
    dTx <- si*I + c*b*Tx - rt*Tx
    # Cumulative incidence 
    dC <- beta*S*I/N
    # return the rates of change as a list
    list(c(dS, dI, dTx, dC))
  })
}
```

We then define parameters and starting compartment sizes for ode solver.
```{r}
#2. Define parameters and starting compartment sizes
parameters <- c(beta = 0.5, #effective contact rate (aka transmission rate)
                rt = 0.1, # 1/duration of treatment and tracing
                si = 0.1, # rate of seeking treatment
                c = 0.05, # rate of contact tracing
                b = 0.05# probability that a traced individual is infected
)
# Initial state
state <- c(S = 99999, #population of 100,000, 1 person starts of infected
             I = 1, 
             Tx = 0,
             C = 0)
```

We set the total time steps and run the SI model. 
```{r}
T_end <- 500 #run model for 500 time steps (e.g. months)
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps (e.g. months), and computes output at each time step 

#Run the base-case
output.ct <- as.data.frame(ode(y = state, times = times, func = SI_CC, parms = parameters))
```

Plot S. I, R compartment over time
```{r, echo=FALSE}
output.ct.t <- melt(output.ct %>% select(-C), id.vars="time")
ggplot(output.ct.t) +
  geom_line(aes(time,value,color=variable))+
  theme_bw()
```

Run sensitivity analysis on contact tracing rate and discuss how it affects the dynamics.
```{r}
# Sensitivity analysis on contact tracing rate
c_list <- c(0,0.3,0.5) # a vector of contact tracing rate
output_dt <- data.frame() # empty data to save outcomes

for (this_c in c_list){
  temp <- parameters
  # Replace latent period with the next value in the last
  temp['c'] <- this_c
  # Run ode solver
  this_output <- data.frame(ode(y = state, times = times, func = SI_CC, parms = temp))
  # Record current value of t_lat
  this_output$c = as.character(this_c)
  # Stack the result 
  output_dt <- rbind(output_dt, this_output)
  
}

# Plot the results with varying latent period
ggplot(output_dt)+
  geom_line(aes(x=time, y=I, color=c, group=c))+
  ylab("Infected")+
  xlab("Time")+
  theme_bw()
```

