---
title: "SIR with heterogeneous mixing"
author: "Tess Ryckman"
date: "2023-10-17"
output: html_document
---

1. Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve) 
library(tidyverse) 
library(cowplot)
```

2. Define parameters, starting compartment sizes, and timesteps
In SIR_basic, we had a beta (effective contact rate) of 0.5
Let's say this represented 5 contacts with a 10% probability of infection per infected contact
Now we have a contact matrix instead - still with a 10% probability of infection per contact, but with heterogeneous contact structure. WAIFW: effective contact rate to h from h, to l from h, to h from l, to l from l (in that order).
```{r}
parameters <- list(contact_matrix=matrix(data=c(29, 1, 1, 5)*0.1, 
                                         nrow=2, ncol=2, byrow=T), 
                   gamma = 0.3 #recovery rate (1/duration infection)
)


state <- c(S_High = 24999, #high (h) and low (l) risk groups
           S_Low = 75000,
           I_High = 1, 
           I_Low = 0,
           R_High = 0,
           R_Low = 0
)

T_end <- 500 #run model for 500 time steps (e.g. months)
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps (e.g. months), and computes output at each time step 


print(parameters)
print(state)
```


3. Define function for a basic SIR model with 2 groups and no demography
We structure "state" as a matrix within the function and use matrix multiplication to increase efficiency
```{r}
MixingSIR<-function(t, state, parameters) {

    state <- matrix(data=state, nrow=ncol(parameters$contact_matrix))
    colnames(state) <- c("S", "I", "R")
    rownames(state) <- c("h", "l")
    
    with(parameters, {
      dS <- -1*state[, "S"]*state[,"I"]%*%contact_matrix/sum(state) 
      dI <- state[, "S"]*state[,"I"]%*%contact_matrix/sum(state) -
        gamma*state[,"I"]
      dR <- gamma*state[,"I"]
      
      #return the rates of change as a list
      list(c(dS, dI, dR)) 
    })
}

#demonstrate matrix multiplication here
state_mat <- matrix(data=state, nrow=ncol(parameters$contact_matrix))
colnames(state_mat) <- c("S", "I", "R")
rownames(state_mat) <- c("h", "l")

print(state_mat)

#rates at which high- and low-risk susceptibles acquire infection
state_mat[,"I"]%*%parameters$contact_matrix/sum(state_mat)
```


4. Run the model and view model output
We define a new plotting function that allows for easy visualization of the 2 risk groups
```{r}
output <- ode(y = state, times = times, func = MixingSIR, parms = parameters)
output <- as.data.frame(output) 

output_long <- pivot_longer(output, cols=2:ncol(output), names_to=c("state", "risk"), names_sep="_", values_to="size")
output_long <- output_long %>% group_by(time, risk) %>% mutate(N=sum(size))

plot_trace_risk <-function(out) {
  fig <- ggplot(out, aes(x=time, y=size/N, color=state, linetype=risk)) + 
    geom_line(linewidth=1.25) +
    labs(x='Time', y='Compartment proportions', color='', linetype='Risk') +
    theme_bw() + theme(panel.grid=element_blank())
  return(fig)
}

plot_trace_risk(output_long %>% filter(time<100))
```

5. Compare output to a situation with no risk stratification (using BasicSIR)
```{r}

#this is just the same as in SIR_basic
parameters_basic <- c(beta = 0.5, #effective contact rate (aka transmission rate)
                      gamma = 0.3 #recovery rate (1/duration of infection)
)

state_basic <- c(S = 99999, #population of 100,000, 1 person starts of infected
                 I = 1, 
                 R = 0
)

BasicSIR<-function(t, state, parameters) {
  with(as.list(c(state, parameters)),{ 
    
    N = S + I + R #define N (total population size)
    
    #SIR model equations from lecture - rates of change in and out of each compartment 
    dS <- -beta*S*I/N
    dI <- beta*S*I/N - gamma*I
    dR <- gamma*I
    
    #return the rates of change as a list
    list(c(dS, dI, dR)) 
  })
}

plot_trace <-function(out) {
  fig <- ggplot(out, aes(x=time, y=size/N, color=state)) + 
    geom_line(linewidth=1.25) +
    labs(x='Time', y='Compartment size', color='') +
    theme_bw() + theme(panel.grid=element_blank())
  return(fig)
}

output_basic <- ode(y = state_basic, times = times, func = BasicSIR, parms = parameters_basic)
output_basic <- as.data.frame(output_basic) %>% mutate(N=S+I+R)
output_long_basic <- pivot_longer(output_basic, cols=c("S","I","R"), names_to="state", values_to="size")
fig1 <- plot_trace(output_long_basic %>% filter(time<100))

#sum across risk groups to produce an equivalent trace plot
output_long_sum <- output_long %>% group_by(time, state) %>% 
  summarise(size=sum(size), N=sum(N))
fig2 <- plot_trace(output_long_sum %>% filter(time<100))

#compare trace plots
plot_grid(fig1 + ggtitle("Homogeneous Mixing"), 
          fig2 + ggtitle("Heterogeneous Mixing"),
          align="hv")
```

6. Probably skip for time/wasn't covered in slides, but we can also compare R0.
R0 is greater in the stratified version (explains earlier epidemic peak)
But Rt (harder to calculate) declines faster in the stratified version - exhaustion of susceptibles in the high-risk group
```{r}
R_matrix <- parameters$contact_matrix*(rowSums(matrix(data=state, nrow=ncol(parameters$contact_matrix)))/sum(state))/
  (parameters$gamma)
R0 <- max(eigen(R_matrix)$values) #this is the R0 of the stratified version

R0_basic <- parameters_basic[["beta"]]/parameters_basic[["gamma"]]
print(paste0("R0 basic: ", round(R0_basic, 2), "; R0 stratified: ", round(R0, 2)))
```


7. Changing mixing patterns - but keeping total # contacts the same
```{r}
contact_matrices <- list(matrix(data=c(35, 0, 0, 5)*0.1,
                             nrow=2, ncol=2, byrow=T),
                      matrix(data=c(29, 1, 1, 5)*0.1, 
                             nrow=2, ncol=2, byrow=T),
                      matrix(data=c(17, 3, 3, 5)*0.1,
                             nrow=2, ncol=2, byrow=T),
                      matrix(data=c(5,5,5,5)*0.1,
                             nrow=2, ncol=2, byrow=T)
                      )

names(contact_matrices) <- c("No mixing", "More assortative mixing", 
                             "Less assortative mixing", "Homogeneous mixing")

output_all <- list()
for(i in names(contact_matrices)) {
    parameters[["contact_matrix"]] <- contact_matrices[[i]]
    output <- ode(y = state, times = times, func = MixingSIR, parms = parameters)
    output_long <- pivot_longer(as.data.frame(output), cols=2:ncol(output), 
                                names_to=c("state", "risk"), names_sep="_", values_to="size")
    output_long <- output_long %>% group_by(time, risk) %>% mutate(N=sum(size))
    output_all <- c(output_all, list(output_long %>% mutate(contact_pattern=i)))
}

#results stratified by risk group
output_all <- bind_rows(output_all) %>%
  mutate(contact_pattern=factor(contact_pattern, levels=names(contact_matrices)))
plot_trace_risk(output_all %>% filter(time<=100)) + facet_wrap(~contact_pattern)

#total population results
output_all_sum <- output_all %>% group_by(time, state, contact_pattern) %>% 
  summarise(size=sum(size), N=sum(N))
plot_trace(output_all_sum %>% filter(time<100)) + facet_wrap(~contact_pattern)
```


8. We can also confirm that the homogeneous mixing version is the same as the BasicSIR model
```{r}
plot_trace(output_all_sum %>% filter(time<200 & contact_pattern=="Homogeneous mixing"))
plot_trace(output_long_basic %>% filter(time<200))
```

