---
title: "SIR_demography_Vax_Quarantine"
author: "Kyu Lee"
date: "2023-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggplot2)
library(reshape2)
library(dplyr)
library(ggpubr)
```

## Adds waning immunity to the basic SIR model
This code is to illustrate how to model vaccination and quarantine in SIR model.

First, let's look at the case where we assume that vaccine provides perfect protection from infection. Here, we defined the ode function with S,I,R compartment. In the differential equation, we have a direct flow from S to R, indicating those who are vaccinated and protected from infection.
```{r}
## A. Vaccination provides perfect protection from infection: S -> R ##
OpenSIR_Vax_pp<-function(t, state, parameters) {
  with(as.list(c(state, parameters)),{
    N = S + I + R
    
    #SIR w/ demography equations from lecture
    dS <- -beta*S*I/N + birth*N - death*S - mu*S
    dI <- beta*S*I/N - death*I - gamma*I
    dR <- gamma*I + mu*S - death*R 
    # cumulative incidence
    dC <- beta*S*I/N
    # return the rates of change as a list
    list(c(dS, dI, dR, dC))
  })
}
```

We then define parameters and starting compartment sizes for ode solver. Mu is vaccination rate.
```{r}
# 2. Define parameters and starting compartment sizes
parameters <- c(beta = 0.5, #effective contact rate (aka transmission rate)
                beta.1 = 0.5, # transmission rate given no-vaccination
                mbeta = 0.5, # this should be lower than 1
                gamma = 0.3, #recovery rate (1/duration infection)
                birth = 0.03, #birth rate (per capita)
                death = 0.03, #all-cause mortality rate
                omega = 0, # waning immunity
                mu = 0.01, # vaccination rate
                alpha = 0.3, # vaccine effectiveness
                q = 0.02 # quarantine rate
)

# Initial state distribution
state.pp <- c(S = 99999, #population of 100,000, 1 person starts of infected
              I = 1, 
              R = 0,
              C = 0)
```

We set the total time steps and run the SIR model with perfect protection from vaccination.
```{r}
T_end <- 500 #run model for 500 time steps (e.g. months)
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps (e.g. months), and computes output at each time step 

# Run the base-case
output.vax.pp <- ode(y = state.pp, times = times, func = OpenSIR_Vax_pp, parms = parameters)
```

Plot the number of individuals in S, I, R compartment over time.
```{r, echo=FALSE}
# Plot the basecase result
plot(output.vax.pp)
```

Let's compare the size of I compartment when mu = 0 (no vax) and mu = 0.01 (vax)
```{r}
# Copy parameters
temp_param <- parameters 
# Set vaccination rate = 0
temp_param['mu'] = 0
# Run ode solver and save the outcome in temperary output variable
temp_output <-  ode(y = state.pp, times = times, func = OpenSIR_Vax_pp, parms = temp_param)
# Create a dataset to compare no_vax and vax output
I_dt <- data.frame(time = temp_output[,'time'],no_vax = temp_output[,'I'], vax = output.vax.pp[,'I'])
# Plot the dataset
ggplot(I_dt)+
  geom_line(aes(time,vax,color='vax'))+
  geom_line(aes(time,no_vax,color='No vax'))+
  theme_bw()+
  scale_color_manual(values=c("blue","tomato"))
```


However, vaccination usually doesn't provide perfect protection from infection. So let's see how we can model imperfect protection with vaccination. Another consideration to add is whether or not those who are infected after vaccination are less infectious than those who are infected without vaccination. So here we start with the model that assumes imperfect protection and no change in infectiousness with vaccination.

Below, we defined the ode function with S,I,R compartment, with stratification by vaccine status. Non-vaccinated compartment moves to the vaccinated corresponding compartment at the rate of mu. In beta*S*I term, note that total I who can infect others is the sum of I_NotV and I_V. (1-alpha) indicates reduced susceptibility for vaccinated susceptibles and the higher alpha is, the lower the susceptibility would be.
```{r}
## B. Vaccination provides imperfect protection from infection ##
# B-1. No change in infectiousness
# 1. Define model function
OpenSIR_Vax_ip1<-function(t, state, parameters) {
  with(as.list(c(state, parameters)),{
    N = S_NotV + S_V + I_NotV + I_V + R_NotV + R_V
    
    #compartments without vaccination
    dS_NotV <- -beta*S_NotV*(I_NotV+I_V)/N + birth*N - death*S_NotV - mu*S_NotV
    dI_NotV <- beta*S_NotV*(I_NotV+I_V)/N - death*I_NotV - gamma*I_NotV
    dR_NotV <- gamma*I_NotV - death*R_NotV 
    #compartments with vaccination
    dS_V <- -beta*S_V*(I_NotV+I_V)/N*(1-alpha) - death*S_V + mu*S_NotV
    dI_V <- beta*S_V*(I_NotV+I_V)/N*(1-alpha) - death*I_V - gamma*I_V
    dR_V <- gamma*I_V - death*R_V 
    
    #cumulative number of cases
    dC <- beta*S_NotV*(I_NotV+I_V)/N + beta*S_V*(I_NotV+I_V)/N*(1-alpha)
    
    # return the rates of change as a list
    list(c(dS_NotV,dS_V,dI_NotV,dI_V, dR_NotV,dR_V,dC))
  })
}
```

And here is the model that assumes imperfect protection and reduced infectiousness with vaccination. Now we apply separate beta values (beta.1 and beta.2) to I_NotV and I_V. Beta.2 is calculated as a fraction of beta.1 (fraction: mbeta).
```{r}
# 1. Define model function
# B-2. Change in risk of infecting others with vaccination
OpenSIR_Vax_ip2<-function(t, state, parameters) {
  with(as.list(c(state, parameters)),{
    N = S_NotV + S_V + I_NotV + I_V + R_NotV + R_V
    beta.2 = mbeta*beta.1 # transmission rate given vaccination
    #compartments without vaccination
    dS_NotV <- -beta.1*S_NotV*I_NotV/N - beta.2*S_NotV*I_V/N + birth*N - death*S_NotV - mu*S_NotV
    dI_NotV <- beta.1*S_NotV*I_NotV/N + beta.2*S_NotV*I_V/N - death*I_NotV - gamma*I_NotV
    dR_NotV <- gamma*I_NotV - death*R_NotV 
    #compartments with vaccination
    dS_V <- -beta.1*S_V*I_NotV/N*(1-alpha) - beta.2*S_V*I_V/N*(1-alpha) - death*S_V + mu*S_NotV
    dI_V <- beta.1*S_V*I_NotV/N*(1-alpha) + beta.2*S_V*I_V/N*(1-alpha) - death*I_V - gamma*I_V
    dR_V <- gamma*I_V - death*R_V 
    
    #cumulative number of cases
    dC <- beta.1*S_NotV*I_NotV/N + beta.1*S_V*I_NotV/N*(1-alpha) + beta.2*S_NotV*I_V/N + beta.2*S_V*I_V/N*(1-alpha)
    
    # return the rates of change as a list
    list(c(dS_NotV,dS_V,dI_NotV,dI_V, dR_NotV,dR_V, dC))
  })
}
```

We then define parameters and starting compartment sizes for ode solver. Mu is vaccination rate. mbeta is set to 1 if no change in infectiousness and a value less than 1 if we assume reduced infectiousness. alpha is vaccine effectiveness.
```{r}
# 2. Define parameters and starting compartment sizes
parameters <- c(beta = 0.5, #effective contact rate (aka transmission rate)
                beta.1 = 0.5, # transmission rate given no-vaccination
                mbeta = 0.5, # 1 if no change in infectiousness with vaccination, 
                           # less than 1 if reduced infectiousness with vaccination
                gamma = 0.3, #recovery rate (1/duration infection)
                birth = 0.03, #birth rate (per capita)
                death = 0.03, #all-cause mortality rate
                omega = 0, # waning immunity
                mu = 0.01, # vaccination rate
                alpha = 0.3, # vaccine effectiveness
                q = 0.02 # quarantine rate
                
)

# initial state
state.ip <- c(S_NotV = 99999,
              S_V = 0,
              I_NotV = 1,
              I_V = 0,
              R_NotV = 0,
              R_V = 0,
              C = 0)
```

We set the total time steps and run the SIR model with imperfect protection from vaccination. Because each compartment is stratified, we computed the total S, I, R compartment
```{r}
T_end <- 500 #run model for 500 time steps (e.g. months)
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps (e.g. months), and computes output at each time step 

# Run ODE solver
output.vax.ip1 <- ode(y = state.ip, times = times, func = OpenSIR_Vax_ip1, parms = parameters)
output.vax.ip2 <- ode(y = state.ip, times = times, func = OpenSIR_Vax_ip2, parms = parameters)
# Calculate S, I, and R
output.vax.ip1 <- as.data.frame(output.vax.ip1) %>%
  mutate(S = S_V + S_NotV,
         I = I_V + I_NotV,
         R = R_V + R_NotV)
output.vax.ip2 <- as.data.frame(output.vax.ip2) %>%
  mutate(S = S_V + S_NotV,
         I = I_V + I_NotV,
         R = R_V + R_NotV)  
```

Plot the S, I, R compartment from the base-case outcomes and compare between the two models.
```{r, echo=FALSE}
# Plot base-case outcomes
g1<- ggplot(output.vax.ip1)+
        geom_line(aes(time,S,color="S"))+
        geom_line(aes(time,I,color="I"))+
        geom_line(aes(time,R,color="R"))+
        scale_color_manual(values=c("tomato","blue","green"))+
        ggtitle("No Change in Infectiousness")+
        ylab("Number of individuals")+
        theme_bw()
g2<- ggplot(output.vax.ip2)+
        geom_line(aes(time,S,color="S"))+
        geom_line(aes(time,I,color="I"))+
        geom_line(aes(time,R,color="R"))+
        scale_color_manual(values=c("tomato","blue","green"))+
        ggtitle("Reduced infectiousness with vaccination")+
        ylab("Number of individuals")+
       theme_bw()
# side-by-side plot
ggarrange(g1,g2,common.legend = TRUE)
```

## Sensitivity Analysis
In the next set of codes, we are going to perform sensitivity analysis on (1) vaccination rate, (2) vaccine effectiveness, and (3) the ratio in effective contact rate between vaccinated and non-vaccinated individuals to see how varying those parameter values changes projected epidemics. (1) and (2) will be done in OpenSIR_Vax_ip1 and (3) will be done in OpenSIR_Vax_ip2.

Discuss how changing vaccination rate changes the projected epidemic.
```{r}
# 1. Vaccination rate (on OpenSIR_Vax_ip1)
mu_list <- seq(0,0.1,length.out=3) # a vector of vaccination rate
mu_out_all <- data.frame() # an empty dataset to save outputs

for(this_mu in mu_list){
  temp_param <- parameters # copy parameters
  temp_param[['mu']] <- this_mu # replace mu 
  this_out <- data.frame(ode(y = state.ip, times = times, func = OpenSIR_Vax_ip1, parms = temp_param)) # run ode solver
  # compute S, I, R and save mu in this cycle
  this_out <- this_out %>%
    mutate(N = S_NotV + S_V + I_NotV + I_V + R_NotV + R_V,
           mu = this_mu,
           S = S_NotV + S_V,#total susceptible
           I = I_NotV + I_V, # total infected
           R = R_NotV + R_V) # total recovered
  # stack the result
  if(this_mu == mu_list[1]){
    mu_out_all <- this_out
  }else{
    mu_out_all <- rbind(mu_out_all, this_out)
  }
}
# Change the output data to long-form data
mu_out_all_t <- melt(mu_out_all %>% select(time,S,I,R,N,mu), id.vars=c("time","mu","N"))
# Plot S, I, R
ggplot(mu_out_all_t) +
  geom_line(aes(x=time,y=value/N*100,color=variable))+
  ylab("Percentage of population")+
  facet_wrap(.~mu)+
  theme_bw()
# Plot I stratified by vaccination status
ggplot(mu_out_all)+
  geom_point(aes(x=time,y=I_NotV,color='Not vaccinated'))+
  geom_point(aes(x=time,y=I_V,color='Vaccinated'))+
  ylab("Infections")+
  facet_wrap(.~mu)+
  scale_color_manual(values=c("green","tomato"))+
  theme_bw()
# Plot prevalence of I (I/N)
ggplot(mu_out_all) + 
  geom_point(aes(x=time,y=I/N))+
  facet_wrap(.~mu)+
  ylab("Proportion of population who are infected")+
  theme_bw()
# Cumulative number of infections
ggplot(mu_out_all) + 
  geom_point(aes(x=time,y=C))+
  facet_wrap(.~mu)+
  ylab("Cumulative infections")+
  theme_bw()
```

How does increasing vaccine effectiveness changes the projected epidemic?
```{r}
# 2. Vaccine effectiveness
alpha_list <- seq(0,0.3,length.out=3) # a vector of vaccine effectiveness
alpha_out_all <- data.frame() # an empty dataset to save outputs

for(this_alpha in alpha_list){
  temp_param <- parameters # copy parameters
  temp_param[['alpha']] <- this_alpha # replace alpha
  this_out <- data.frame(ode(y = state.ip, times = times, func = OpenSIR_Vax_ip1, parms = temp_param)) # run ode solver
  # compute S, I, R and save mu in this cycle
  this_out <- this_out %>%
    mutate(N = S_NotV + S_V + I_NotV + I_V + R_NotV + R_V,
           alpha = this_alpha,
           S = S_NotV + S_V,
           I = I_NotV + I_V,
           R = R_NotV + R_V)
  # Stack the result
  if(this_alpha == alpha_list[1]){
    alpha_out_all <- this_out
  }else{
    alpha_out_all <- rbind(alpha_out_all, this_out)
  }
}
# Change the outcome data to long-form data
alpha_out_all_t <- melt(alpha_out_all%>% select(time,S,I,R,N,alpha), id.vars=c("time","alpha","N"))
# Plot S, I, R
ggplot(alpha_out_all_t) +
  geom_line(aes(x=time,y=value/N*100,color=variable))+
  ylab("Percentage of population")+
  facet_wrap(.~alpha)+
  theme_bw()
# Plot S, I, R stratified by vaccination status
ggplot(alpha_out_all)+
  geom_point(aes(x=time,y=I_NotV,color='Not vaccinated'))+
  geom_point(aes(x=time,y=I_V,color='Vaccinated'))+
  ylab("Infections")+
  facet_wrap(.~alpha)+
  scale_color_manual(values=c("green","tomato"))+
  theme_bw()
# Plot prevalence of I
ggplot(alpha_out_all) + 
  geom_point(aes(x=time,y=I/N))+
  facet_wrap(.~alpha)+
  ylab("Prevalence of infection")+
  theme_bw()
# Plot cumulative number of infections
ggplot(alpha_out_all) + 
  geom_point(aes(x=time,y=C))+
  facet_wrap(.~alpha)+
  ylab("Cumulative infections")+
  theme_bw()
# Breakthrough infection
ggplot(alpha_out_all)+
  geom_point(aes(x=time,y=I_V))+
  facet_wrap(.~alpha)+
  ylab("Number of breakthrough infections")+
  theme_bw()
```

How does the projected epidemic change as we decrease the ratio of effective contact rate?
```{r}
# 3. Ratio of effective contact rate among infected individuals with and without vaccination 
mbeta_list <- seq(1,0.7,length.out=3) # a vector of mbeta
mbeta_out_all <- data.frame() # an empty dataset to save outputs

for(this_mbeta in mbeta_list){
  temp_param <- parameters # copy parameters
  temp_param[['mbeta']] <- this_mbeta # replace mbeta
  this_out <- data.frame(ode(y = state.ip, times = times, func = OpenSIR_Vax_ip2, parms = temp_param)) # run ode solver
  # compute S,I,R and save mbeta in this cycle
  this_out <- this_out %>%
    mutate(N = S_NotV + S_V + I_NotV + I_V + R_NotV + R_V,
           mbeta = this_mbeta,
           S = S_NotV + S_V,
           I = I_NotV + I_V,
           R = R_NotV + R_V)
  # Stack the result
  if(this_mbeta == mbeta_list[1]){
    mbeta_out_all <- this_out
  }else{
    mbeta_out_all <- rbind(mbeta_out_all, this_out)
  }
}
# change the data to long-form data
mbeta_out_all_t <- melt(mbeta_out_all%>% select(time,S,I,R,N,mbeta), id.vars=c("time","mbeta","N"))
# Plot S,I,R
ggplot(mbeta_out_all_t) +
  geom_line(aes(x=time,y=value/N*100,color=variable))+
  facet_wrap(.~mbeta)+
  ylab("Proportion of population")+
  theme_bw()
# plot cumulative number of infections
ggplot(mbeta_out_all) + 
  geom_point(aes(x=time,y=C))+
  facet_wrap(.~mbeta)+
  ylab("Cumulative infections")+
  theme_bw()
```

Let's move on to modeling quarantine in SIR. In the differential equation, we have Q compartment that indicates individuals who are infected and isolated under quarantine. 

```{R}
# C. SIR model with quarantine
OpenSIR_Qrtn<-function(t, state, parameters) {
  with(as.list(c(state, parameters)),{
    N = S + I + Q + R 
    
    #SIR with quarantine
    dS<- -beta*S*I/N + birth*N - death*S
    dI <- beta*S*I/N - death*I - q*I - gamma*I
    dQ <- q*I - death*Q - gamma*Q
    dR <- gamma*I + gamma*Q - death*R
    dC <- beta*S*I/N
    
    # return the rates of change as a list
    list(c(dS, dI, dQ, dR, dC))
  })
}
```

We then define parameters and starting compartment sizes for ode solver. q is quarantine rate (I->Q)
```{r}
# 2. Define parameters and starting compartment sizes
parameters <- c(beta = 0.5, #effective contact rate (aka transmission rate)
                beta.1 = 0.5, # transmission rate given no-vaccination
                mbeta = 0.5, # 1 if no change in infectiousness with vaccination, 
                           # less than 1 if reduced infectiousness with vaccination
                gamma = 0.3, #recovery rate (1/duration infection)
                birth = 0.03, #birth rate (per capita)
                death = 0.03, #all-cause mortality rate
                omega = 0, # waning immunity
                mu = 0.01, # vaccination rate
                alpha = 0.3, # vaccine effectiveness
                q = 0.02 # quarantine rate
                
)
# Initial state
state.q <- c(S = 99999, #population of 100,000, 1 person starts of infected
             I = 1, 
             Q = 0,
             R = 0,
             C = 0)
```

We set the total time steps and run the SIR model with quarantine
```{r}
T_end <- 500 #run model for 500 time steps (e.g. months)
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps (e.g. months), and computes output at each time step 
# Run ODE solver
output.qrtn <- ode(y = state.q, times = times, func = OpenSIR_Qrtn, parms = parameters)
```

Plot S,I,R,Q compartment
```{r, echo=FALSE}
# Plot base-case result
output.qrtn.t <- melt(as.data.frame(output.qrtn) %>% select(-C), id.vars="time")
ggplot(output.qrtn.t)+
  geom_line(aes(time,value,color=variable))+
  ylab("Number of individuals")+
  theme_bw()
```

Let's run a sensitivity analysis on quarnatine rate to see how it affects the projected epidemic.
```{r}
# 4. Quarantine rate
q_list <- c(0,0.05,0.1) # a vector of quarantine rate
q_out_all <- data.frame() # an empty dataset to save the outcome

for(this_q in q_list){
  temp_param <- parameters #copy parameter
  temp_param[['q']] <- this_q # replace q
  # run ode solver
  this_out <- data.frame(ode(y = state.q, times = times, func = OpenSIR_Qrtn, parms = temp_param))
  # save q value in this cycle
  this_out <- this_out %>%
    mutate(N = S+I+R+Q,
           q = this_q)  
  # stack the result
  if(this_q == q_list[1]){
    q_out_all <- this_out
  }else{
    q_out_all <- rbind(q_out_all, this_out)
  }
}
# change the dataset to a long-form dataset
q_out_all_t <- melt(q_out_all%>% select(time,S,I,R,N,q), id.vars=c("time","q","N"))
# Plot S, I, R
ggplot(q_out_all_t) +
  geom_line(aes(x=time,y=value/N*100,color=variable))+
  ylab("Proportion of population")+
  facet_wrap(.~q)+
  theme_bw()
# Plot cumulative number of infections
ggplot(q_out_all) + 
  geom_point(aes(x=time,y=C))+
  facet_wrap(.~q)+
  ylab("Cumulative infections")+
  theme_bw()
```

