---
title: "SIRS_waningimmunity"
author: "Kyu Lee"
date: "2023-10-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggplot2)
library(reshape2)
```

## Adds waning immunity to the basic SIR model
This code is to illustrate how to model waning immunity in SIR model.

First, we defined the ode function with S,I,R compartment. In the differential equation, we have a backward flow from R to S, indicating those who lost infection-induced immunity over time.
```{r}
#1. Define model function
OpenSIRS<-function(t, state, parameters) {
  with(as.list(c(state, parameters)),{
    N = S + I + R
    
    #SIR w/ demography equations from lecture
    dS <- -beta*S*I/N + birth*N - death*S + omega*R
    dI <- beta*S*I/N - death*I - gamma*I
    dR <- gamma*I - death*R - omega*R
    
    # return the rates of change as a list
    list(c(dS, dI, dR))
  })
}
```

We then define parameters and starting compartment sizes for ode solver. Omega is 1/duration of immunity, assuming a constant rate of waning immunity.
```{r}
#2. Define parameters and starting compartment sizes
parameters <- c(beta = 0.5, #effective contact rate (aka transmission rate)
                gamma = 0.3, #recovery rate (1/duration infection)
                birth = 0.03, #birth rate (per capita)
                death = 0.03, #all-cause mortality rate
                omega = 0.3 # waning immunity
)

state <- c(S = 99999, #population of 100,000, 1 person starts of infected
           I = 1, 
           R = 0
)
```

We set the total time steps and run the SIRS model. 
```{r}
#3. Run SIRS model 
T_end <- 500 #run model for 500 time steps (e.g. months)
times <- seq(0, T_end, by = 1) #runs the model for 500 time steps (e.g. months), and computes output at each time step 
output <- ode(y = state, times = times, func = OpenSIRS, parms = parameters)
```

Plot the number of individuals in S, I, R compartment over time.
```{r plot, echo=FALSE}
output_t <- melt(as.data.frame(output), id.vars='time')
ggplot(output_t)+
  geom_line(aes(time,value,color=variable))+
  theme_bw()
```

Let's look at how the dynamic changes as we vary rate of waning immunity
```{r}
#Run one-way sensitivity analysis on the rate of waning immunity
omega_list <- seq(0,0.3,by=0.1) # a vector of omegas
osa_dt <- data.frame() # empty data to save outputs

for (this_omega in omega_list){
  # Update omega
  parameters['omega']= this_omega
  # Solve ODE
  this_output <- as.data.frame(ode(y = state, times=times, func=OpenSIRS, parms = parameters))
  # Save omega value in the outcome table
  this_output$omega = as.character(this_omega)
  # Stack the ode result
  osa_dt <- rbind(osa_dt, this_output)
  
}
```

Plot the one-way sensitivity analysis. How does changing waning immunity affect the disease dynamic?
```{r, echo=FALSE}
# Plot the results with varying rate of waning immunity
ggplot(osa_dt)+
    geom_line(aes(x=time, y=I, color=omega, group=omega))+
    ylab("Infected")+
    xlab("Time")+
    theme_bw()
```
Effect of waning immunity on epidemic growth can depend on how fast the disease spreads(beta). Let's vary both waning immunity (omega) and effective contact rate(beta) in two-way sensitivity analysis.
```{r}
#Run two-way sensitivity analysis on omega & beta
omega_list <- seq(0,0.3,by=0.1) # a vector of omegas
beta_list <- seq(0.4,0.6,by=0.1) # a vector of beta values
twsa_dt <- data.frame() # data to save the ode outcomes

for (this_beta in beta_list){
  # Update beta
  parameters['beta'] = this_beta
  for (this_omega in omega_list){
    # update omega
    parameters['omega'] = this_omega
    # Run ode solver
    this_output <- as.data.frame(ode(y = state, times=times, func=OpenSIRS, parms = parameters))
    # Save omega
    this_output$omega = as.character(this_omega)
    # Save beta
    this_output$beta = as.character(this_beta)
    # Stack the ode result
    twsa_dt <- rbind(twsa_dt, this_output)
  }
}
```

Plot the results with varying omega and beta. Has the effect of waning immunity on epidemic changed with different betas?
```{r, echo=FALSE}
# Change the label for omega and beta
omega_lb <- sapply(omega_list,function(x) paste0("omega=",x))
omega_lb
twsa_dt$omega <- factor(twsa_dt$omega, labels = omega_lb)
beta_lb <- sapply(beta_list,function(x) paste0("beta=",x))
beta_lb
twsa_dt$beta <- factor(twsa_dt$beta, labels = beta_lb)
# Check if label is created correctly
head(twsa_dt)

# Plot the two-way sensitivity analysis result
ggplot(twsa_dt)+
  geom_line(aes(x=time, y=I))+
  facet_grid(beta~omega)+
  ylab("Infected")+
  xlab("Time")+
  theme_bw()
```